A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-04-14 manuscript.qmd
add rcode.qmd notebook and experiment which equal-contributor yaml option (seems currently bugged)

---
title: "A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models"
shorttitle: "Simulation-Based Sample Size Planning for GLMMs"
author:
  - name: Timo K. Koch
    corresponding: true
    orcid: 0000-0001-6728-2063
    email: timo.koch@unisg.ch
    roles:
      - conceptualization
      - formal analysis
      - writing
      - editing
    affiliations:
      - name: University of St. Gallen 
        department: Institute of Behavioral Science & Technology
        address: Torstrasse 25
        city: St. Gallen
        country: Switzerland
        postal-code: 9000
      - id: "lmu"
        name: LMU Munich
        department: Department of Psychology
  - name: Florian Pargent
    orcid: 0000-0002-2388-553X
    roles:
      - formal analysis
      - methodology
      - visualization
      - writing
      - editing
    affiliations:
      - ref: lmu
  - name: Anne-Kathrin Kleine
    roles:
      - formal analysis
      - editing
    affiliations:
      - ref: lmu
  - name: Eva Lermer
    roles:
      - conceptualization
      - funding aquisition
      - editing
    affiliations:
      - ref: lmu
      - name: Technical University of Applied Sciences Augsburg
        department: Department of Business Psychology
  - name: Susanne Gaube
    roles:
      - conceptualization
      - supervision
      - editing
    affiliations:
      - ref: lmu
      - name: University College London
        department: Global Business School for Health
author-note:
  disclosures:
    data-sharing: "All materials (reproducible manuscript incl. R code, simulation results) are available in the project's repository on the Open Science Framework (OSF) at https://osf.io/dhwf4/"
    conflict-of-interest: "None."
    financial-support: "This research was funded by a grant from the Volkswagen Foundation (Grant No. 98525)."
    gratitude: ~
    authorship-agreements: "Timo K. Koch and Florian Pargent contributed equally to this work."
abstract: "When planning experimental research, determining an appropriate sample size and using suitable statistical models are crucial for robust and informative results. However, the recent replication crisis underlines the need for more rigorous statistical methodology and well-powered designs. Generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. Yet, available methods and software for a priori power analyses for GLMMs are often limited to specific designs, while data simulation approaches offer more flexibility. Based on a practical case study, the current tutorial equips researchers with a step-by-step guide and corresponding code for conducting tailored a priori power analyses to determine appropriate sample sizes with GLMMs. Finally, we give an outlook on the increasing importance of simulation-based power analysis in experimental research."
keywords: [power analysis, data simulation, sample size, generalized linear mixed model]
bibliography: [Simulation Tutorial Paper.bib, r-references.bib]
floatsintext: true
numbered-lines: false
csl: apa.csl
---

@@

```{r}
#| label: probtable
#| echo: false
probtable <- data.frame(
  col1 = 
    c("no advice, no expert", "no advice, expert", "false advice, no expert", 
      "false advice, expert", "correct advice, no expert", "correct advice, expert"),
  col2 =
    c("0.70", "0.90", "0.40", "0.85", "0.90", "0.95"),
  col3 = 
    c("$logit(0.70) = \\beta_0$",
      "$logit(0.90) = \\beta_0 + \\beta_e$", 
      "$logit(0.40) = \\beta_0 + \\beta_a$",
      "$logit(0.85) = \\beta_0 + \\beta_e + \\beta_{a} + \\beta_{ea}$",
      "$logit(0.90) = \\beta_0 + \\beta_a + \\beta_c$",
      "$logit(0.95) = \\beta_0 + \\beta_e + \\beta_a + \\beta_c + \\beta_{ea} + \\beta_{ec}$"))
apa_table(
  probtable
  , col.names = c("Experimental condition",
    "$P(Y=1|\\mathbf{X=x}, \\mathbf{U} = \\mathbf{0})$",
    "Implied equation")
  , caption = "Assumed probabilities that an average subject solves an average item in each experimental condition."
  , note = "Implied equations are derived based on the model equations and setting all random intercept terms to 0."
  , escape = FALSE
)
```

Table \@ref(tab:probtable) shows our set of assumptions concerning the probability that an average subject solves an average item for each experimental condition, as well as the corresponding equations implied by the model.
The table can be used to compute the implied values for the $\beta$ parameters, starting with the first equation and reinserting the computed $\beta$ values in all following equations:

@@

```{r}
pred_df <- expand_grid(advice_present = 0:1, advice_correct = 0:1, expert = 0:1)
pred_df
probs <- predictions(fit, newdata = pred_df, type = "response", re.form = NA)
probs
```

>>>>

---
title: "A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models"
shorttitle: "Simulation-Based Sample Size Planning for GLMMs"
author:
  - name: Timo K. Koch
    corresponding: true
    equal-contributor: true
    orcid: 0000-0001-6728-2063
    email: timo.koch@unisg.ch
    roles:
      - conceptualization
      - formal analysis
      - writing
      - editing
    affiliations:
      - name: University of St. Gallen 
        department: Institute of Behavioral Science & Technology
        address: Torstrasse 25
        city: St. Gallen
        country: Switzerland
        postal-code: 9000
      - id: "lmu"
        name: LMU Munich
        department: Department of Psychology
  - name: Florian Pargent
    equal-contributor: true
    orcid: 0000-0002-2388-553X
    roles:
      - formal analysis
      - methodology
      - visualization
      - writing
      - editing
    affiliations:
      - ref: lmu
  - name: Anne-Kathrin Kleine
    roles:
      - formal analysis
      - editing
    affiliations:
      - ref: lmu
  - name: Eva Lermer
    roles:
      - conceptualization
      - funding aquisition
      - editing
    affiliations:
      - ref: lmu
      - name: Technical University of Applied Sciences Augsburg
        department: Department of Business Psychology
  - name: Susanne Gaube
    roles:
      - conceptualization
      - supervision
      - editing
    affiliations:
      - ref: lmu
      - name: University College London
        department: Global Business School for Health
author-note:
  disclosures:
    data-sharing: "All materials (reproducible manuscript incl. R code, simulation results) are available in the project's repository on the Open Science Framework (OSF) at https://osf.io/dhwf4/"
    conflict-of-interest: "The authors declare that there were no conflicts of interest with respect to the authorship or the publication of this article."
    financial-support: "This research was funded by a grant from the Volkswagen Foundation (Grant No. 98525)."
    gratitude: ~
    authorship-agreements: "Timo K. Koch and Florian Pargent contributed equally to this work."
abstract: "When planning experimental research, determining an appropriate sample size and using suitable statistical models are crucial for robust and informative results. However, the recent replication crisis underlines the need for more rigorous statistical methodology and well-powered designs. Generalized linear mixed models (GLMMs) offer a flexible statistical framework to analyze experimental data with complex (e.g., dependent and hierarchical) data structures. Yet, available methods and software for a priori power analyses for GLMMs are often limited to specific designs, while data simulation approaches offer more flexibility. Based on a practical case study, the current tutorial equips researchers with a step-by-step guide and corresponding code for conducting tailored a priori power analyses to determine appropriate sample sizes with GLMMs. Finally, we give an outlook on the increasing importance of simulation-based power analysis in experimental research."
keywords: [power analysis, data simulation, sample size, generalized linear mixed model]
bibliography: [Simulation Tutorial Paper.bib, r-references.bib]
floatsintext: true
numbered-lines: false
csl: apa.csl
---

@@

```{r}
#| label: tab-probtable
#| echo: false
probtable <- data.frame(
  col1 = 
    c("no advice, no expert", "no advice, expert", "false advice, no expert", 
      "false advice, expert", "correct advice, no expert", "correct advice, expert"),
  col2 =
    c("0.70", "0.90", "0.40", "0.85", "0.90", "0.95"),
  col3 = 
    c("$logit(0.70) = \\beta_0$",
      "$logit(0.90) = \\beta_0 + \\beta_e$", 
      "$logit(0.40) = \\beta_0 + \\beta_a$",
      "$logit(0.85) = \\beta_0 + \\beta_e + \\beta_{a} + \\beta_{ea}$",
      "$logit(0.90) = \\beta_0 + \\beta_a + \\beta_c$",
      "$logit(0.95) = \\beta_0 + \\beta_e + \\beta_a + \\beta_c + \\beta_{ea} + \\beta_{ec}$"))
apa_table(
  probtable
  , col.names = c("Experimental condition",
    "$P(Y=1|\\mathbf{X=x}, \\mathbf{U} = \\mathbf{0})$",
    "Implied equation")
  , caption = "Assumed probabilities that an average subject solves an average item in each experimental condition."
  , note = "Implied equations are derived based on the model equations and setting all random intercept terms to 0."
  , escape = FALSE
)
```

@tab-probtable shows our set of assumptions concerning the probability that an average subject solves an average item for each experimental condition, as well as the corresponding equations implied by the model.
The table can be used to compute the implied values for the $\beta$ parameters, starting with the first equation and reinserting the computed $\beta$ values in all following equations:

@@

```{r}
library(tidyverse)
library(marginaleffects)
pred_df <- expand_grid(advice_present = 0:1, advice_correct = 0:1, expert = 0:1)
pred_df
probs <- predictions(fit, newdata = pred_df, type = "response", re.form = NA)
probs
```
